Preregistration: Changes

### 2022-07-25 index.Rmd
- major updates for revision - included multiple imputation - changed theory - include new data-set - extended discussion

---
title: "Analyzing the effects of COVID-19 related social media use on well-being"
---

On this website, you can find the companion material for the paper _Analyzing the effects of COVID-19 related social media use on well-being_.

>>>>

---
title: "The effects of COVID-19 related social media use on well-being"
---

On this website, you can find the companion material for the paper _The effects of COVID-19 related social media use on well-being_.

### 2022-07-25 manuscript.Rmd
- major updates for revision - included multiple imputation - changed theory - include new data-set - extended discussion

---
title             : "Analyzing the effects of COVID-19 related social media use on well-being"
shorttitle        : "COVID-19 related social media use and well-being"

@@

abstract          : "In times of crisis such as the Corona pandemic citizens need to stay informed about recent events, political decisions, or mandatory protection measures. To this end, many people use various types of media, and increasingly social media. However, because social media are particularly engaging, some find it hard to disconnect. In this preregistered study, I investigated whether using social media for COVID-19 related reasons affected psychological well-being. To answer this question I analyzed data from the Austrian Corona Panel Project, which consists of 3,018 participants. Well-being was measured at all 24 waves, and communication at five specific waves. I ran three random effects within between models, controlling for several stable and varying confounders. Results showed that the effects of COVID-19 related social media use on well-being were very small, arguably too small to matter. Fears that social media use during times of crisis critically impairs well-being are not supported."
keywords          : "COVID-19, well-being, social media, news use, panel study."

@@

documentclass     : "apa6"
classoption       : "man"
output            :
  papaja::apa6_pdf:
    latex_engine  : xelatex
# output            : papaja::apa6_docx
header-includes:
  - \usepackage{endnotes}
  - \setlength{\parskip}{0em}
  - \raggedbottom
  - \note{\clearpage}
---

```{r analysis-preferences, include=F, cache=F}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed,
                      cache=T, echo=F)
```

@@

# load packages
packages <- c("devtools", "english", "ggplot2", "gridExtra", "kableExtra", 
              "lavaan", "lme4", "lmerTest",
              "magrittr", "semPlot", "tidyverse")
lapply(c(packages, "papaja", "td"), library, character.only = TRUE)

# load workspace created from analyses.rmd to use results
load("data/workspace_2.RData")
```

@@

As a result, with this study I want to build on this research and investigate whether or not COVID-19 related social media use affected well-being during the pandemic.
To this end, I analyzed a large-scale panel study from the Austrian Corona Panel Project [@kittelAustrianCoronaPanel2020]. 
The panel consists of 24 waves and has an overall sample size of `r nrow(d_wide) %>% prettyNum(big.mark = ",")` participants.
<!-- At least 1,500 participants took part per wave, and the sample is representative of the Austrian population. -->
The panel study collected a large number of psychological and demographic variables.
<!-- Being able to control for many confounding third variables, both stable and varying, together with the longitudinal design, this creates a unique opportunity  -->
I explicitly aimed to investigate the causal effects of COVID-19 related social media use on well-being.

@@

The hierarchical taxonomy of computer-mediated communication differentiates six levels of how people engage with digital technology.
First, the device (e.g., smartphone); second, the type of application (e.g., social networking site); third, the branded application (e.g., Twitter); fourth, the feature (e.g., status post); fifth, the interaction (e.g., one-to-many); and sixth, the message (e.g., content) [@meierComputermediatedCommunicationSocial2020a].
Whereas the first four levels focus on the communication _channel_, the last two address the communication _type_.
To measure social media use for the consumption of COVID-19 related news and topics, I here employ both the channel and the type of communication perspective, which together should offer a nuanced understanding of communication.

First, I investigate how well-being is affected by different types of communication affect, namely active and passive use.
Defining what constitutes active and what passive use is not always clear, and different understandings are currently discussed [@ellisonWhyWeDon2020; @meierComputermediatedCommunicationSocial2020a].
Reading is generally considered as passive and writing as active, while there are also specific behaviors such as such liking or sharing content that fall somewhere in-between [@meierDoesPassiveSocial2022].
In this study, I hence distinguish (a) _reading_ (passive), (b) _posting_ (active), and (c) _liking and sharing_ COVID-19 related posts (both active and passive).
Second, I analyze how using the most prominent branded applications affects well-being, and whether this effect changes across applications.
Branded apps are separate entities with potentially divergent effects. 
Twitter might have a different effect as compared to WhatsApp because of their respective affordances.
For example, @waterlooNormsOnlineExpressions2018 found that it's more adequate to express negative emotions on WhatsApp than on Twitter or on Instagram. 
The branded applications investigated here are Facebook, Twitter, Instagram, WhatsApp, and YouTube.
Worth noting, this study is not about _general_ social media use during times of COVID, but on social media use _focused_ on COVID-19 related content. 
Examples of such media use include posting thoughts about the pandemic or retweeting COVID-19 related news.

@@

Can media use be such a factor?
In advance, there doesn't seem to be a clear winner, and it's likely that both positive and negative effects cancel each other out.
Empirically, social media use on average oes not have a strong effect on well-being [@meierComputermediatedCommunicationSocial2020a].
According to the Differential Susceptibility to Media Effects Model [@valkenburgDifferentialSusceptibilityMedia2013], the effects of media use differ across individuals. 
Whereas for some media are beneficial, for others they are harmful.
On average, however, effects are often small or negligible.
But also when focusing on individual users, social media can have both positive and negative effects on well-being [@buchiDigitalWellbeingTheory2021].
They can impair well-being when causing embarrassment, stress, or disinformation, and they can improve well-being when providing connectedness, information, or entertainment [@buchiDigitalWellbeingTheory2021].

Two of the most prominent media effect theories argue against strong negative impacts.
First, uses and gratifications theory states that people explicitly and rationally chose specific media because of their respective benefits [@katzUsesGratificationsResearch1973]. 
If those benefits don't exist, they will spend their time elsewhere.
And social media offer copiuos benefits. 
Most prominently, they help find relevant information, maintain and foster relationships, express one's personality, and entertain oneself [@pelletierOneSizeDoesn2020].
The second major theory that needs to be listed here is mood management theory [@zillmannMoodManagementCommunication1988].
Users implicitly learn what media help them balance their mood and affect.
For example, when bored people use social media to entertain themselves.
Using experience sampling of well-being and logs of social media use, a study with 82 participants from Italy found that after episodes of social media use, levels of positive affect increased significantly [@marcianoDynamicsAdolescentsSmartphone2022].

@@

Because media effects can differ across situations [@valkenburgDifferentialSusceptibilityMedia2013], I now briefly focus on the effects of COVID-19 related social media use specifically.
First, one could assume a _direct_ negative effect on well-being, and especially on positive or negative affect, which are more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Consuming such news can be depressing [@dornemannHowGoodBad2021], perhaps even changing some general perspectives on life.
<!-- So, just like being hit by a hammer hurts and we don't need any "mediating mechanism", this could be the case here as well. -->
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects [@dornemannHowGoodBad2021].
A study with 2.057 respondents from Italy reported that during the pandemic virtual community and social connectedness even increased during the pandemic [@guazziniSecondWaveAnalysis2022].
<!-- However, in light of a worldwide pandemic with millions of deaths, the negative direct effect seems more plausible. -->

There could also be _indirect_ effects.
When browsing social media for Covid-19 related news, many users reported being captivated to such an extent that they could not stop using social media [@kleinDarklySoothingCompulsion2021].
During the pandemic social media use was at an all-time high in the US [@statistaAverageDailyTime2021].
Although it is most likely that moderate social media use is not detrimental [@orbenTeenagersScreensSocial2020], overuse, however, might be more critical, and several studies have shown more pronounced negative effects for extreme users [@przybylskiLargescaleTestGoldilocks2017].
To explain, overuse could impair well-being if it replaces meaningful or functional activities such as meeting others, working, actively relaxing, or exercising.
So if a society collectively overuses social media during a pandemic, there might be potential for negative effects.
<!-- If overuse replaces such activities it's reasonable to assume that it's also detrimental. -->
On the other hand, one can make the case that overuse might even be beneficial, especially in times of a pandemic.
Exchanging COVID-19 related messages with friends via WhatsApp might replace the in-person contact one would have otherwise, but which is literally impossible at the time.
In situations where meaningful and functional activities are prohibited, using social media to exchange about COVID-19 related topics might not be the worst idea.
Besides, given that nowadays a large number of experts, scientists, and politicians converse directly on social media, one can get first-hand high quality information on current developments.
<!-- On the other hand, there is of course also much disinformation, and "bingeing" on COVID-19 fake news might also pose risks for impaired well-being. -->

@@

In conclusion, in light of the theoretical considerations and empirical studies presented above, I expect that COVID-19 related communication on social media doesn't affect well-being in a meaningful or relevant way.

@@

The hypothesis explicitly states a causal effect.
In non-experimental studies, longitudinal designs can help investigate causality.
Using longitudinal designs alone, however, is not sufficient for establishing correct causal statements [@rohrerTheseAreNot2021].
In addition, we for example also need to control for confounding third variables, and importantly also for _varying_ third variables.

@@

Thus, only when controlling for _all_ relevant confounders, we can correctly estimate causality without bias [@rohrerThinkingClearlyCorrelations2018].
Obviously, we can never be entirely sure to have included all confounders, which makes absolute statements regarding causality virtually impossible.
In addition, when determining the overall causal effect, we need to make sure _not_ to control for mediating variables [@rohrerThinkingClearlyCorrelations2018], for doing so would bias our assessment of the causal effect.
Complicating matters further, it is often unclear if a variable is a mediator or a confounder.[^collider]
However, despite all these caveats, when controlling for relevant variables (that aren't mediators), we can be much more certain that we measured causality correctly.
The aim should therefore be to collect as many varying and nonvarying confounders as possible (which I believe is seldom done in our field), while knowing that absolute certainty regarding causality cannot be reached.

@@

[^explore]: The dataset includes many other variables that one could also potentially control for, and I invite interested readers to download the and explore potential interesting relationships.

@@

Finally, one precondition of causality is temporal order. 
The cause needs to precede the effect.
Finding the right interval between cause and effect is crucial.
For example, if we want to understand the effect of alcohol consumption on driving performance, it makes a big difference if driving performance is measured one minute, one hour, one day, or one week after consumption.
<!-- Finding the right interval is difficult. -->
If variables are stable, longer intervals are needed; if they fluctuate, shorter intervals.
In the case of well-being, we need shorter intervals for affect and longer ones for life satisfaction [@dienlinImpactDigitalTechnology2020].
Still, choosing the right interval is challenging, because especially short intervals are hard to implement in practice and often require advanced methods such as experience sampling (also known as in situ measurement or ambulant assessment) [@schnauber-stockmannMobileDevicesTools2020].

In this study, I hence analyze how using social media during the last week affected their positive and negative affect during the same week.
More precisely, if people during the last week used COVID-19 related social media more than the usually do, did they feel better or worse during that week than they usually do?
Regarding life satisfaction, I implemented a longer interval.
Namely, if people during the last week used COVID-19 related social media more than the usually do, were they at the end of the week more or less satisfied with their lives than they usually are?
I hence analyze if when a person changes their social media diet, are there (a) _simultaneous_ changes in their affect and (b) _subsequent_ changes in their life satisfaction?
When additionally controlling for both stable and varying confounders, we can then be more sure that the effect is indeed causal.
Similar approaches were implemented by other studies [@johannesNoEffectDifferent2022; @scharkowHowSocialNetwork2020], and they are considered a best practice approach toward analyzing causality [@bellFixedRandomEffects2019].

@@

The data come from the Austrian Corona Panel Project [@kittelAustrianCoronaPanel2021].
The study was conducted between March 2020 and October 2021.
It contains 26 waves, and at the time of writing the first 24 waves were available for download.
Between March 2020 and July 2020, the intervals were weekly, and afterward the intervals were monthly.
Each wave consists of at least 1,500 respondents.
The overall sample size was _N_ = `r d_raw %>% nrow() %>% prettyNum(big.mark = ",")`, and `r d_long %>% nrow() %>% prettyNum(big.mark = ",")` observations were collected.
Panel mortality was compensated through a continuous acquisition of new participants. 
All respondents needed to have access to the internet (via computer or mobile devices such as smartphones or tablets).
They were sampled from a pre-existing online access panel provided by the company Marketagent, Austria.
Respondents were asked and incentivized with 180 credit points to participate in each wave of the panel.

@@

The factorial validity of the scales were tested with confirmatory factor analyses (CFA).
Because Mardia’s test showed that the assumption of multivariate normality was violated, I used the more robust Satorra-Bentler scaled and mean-adjusted test statistic (MLM) as estimator.
To avoid overfitting, I tested the scales on more liberal fit criteria (CFI > .90, TLI > .90, RMSEA <. .10, SRMR < .10) [@klinePrinciplesPracticeStructural2016].
Finally, REWB-models cannot model latent variables.
To increase precision, I therefore exported factor scores from the CFAs for positive and negative affect.
Respondents who answered less than 50% of all questions were removed.
The remaining missing responses were imputed using predictive mean matching.

@@

Social media use was measured for all participants on waves 1, 2, 8, 17, and 23. 
Freshly recruited respondents always answered all questions on social media use.

@@

First, when looking at the variables from a descriptive perspective (Figure \@ref(fig:descriptives)), we see that all well-being measures did not change substantially across the different waves of data collection.
COVID-19 related media use, however, decreased slightly at the beginning of the study and remained stable after approximately six waves.
The initial decrease might be explained by the fact that the collection of data began at the end of March 2020, hence approximately three months after the pandemic began.
It could be that after an initial uptick, COVID-19 related social media use was already declining at the time, returning to more normal levels.

@@

The study's main hypothesis was that the effects of social media use on well-being would be trivial.
Regarding the effects of different communication _types_---that is, reading vs. sharing vs. posting---all within-person effects fell completely within the a-priori defined null region (see Figure \@ref(fig:res-activity)).
For example, respondents who used social media more frequently than usual to read about COVID-19 related topics did not show a simultaneous change in life satisfaction (_b_ = `r dat_fig_results_activity %>% filter(dv == "Life satisfaction" & iv == "Reading" & type == "within" & Analysis == "1. regular") %>% select(estimate) %>% round(2)` [95% CI `r dat_fig_results_activity %>% filter(dv == "Life satisfaction" & iv == "Reading" & type == "within" & Analysis == "1. regular") %>% select(conf.low) %>% round(2)`, `r dat_fig_results_activity %>% filter(dv == "Life satisfaction" & iv == "Reading" & type == "within" & Analysis == "1. regular") %>% select(conf.high) %>% round(2)`]).
All confidence intervals included zero; hence, all effects were also statistically non-significant.
As a result, the hypothesis was supported for all COVID-19 related types of social media communication.

>>>>

---
title             : "The effects of COVID-19 related social media use on well-being"
shorttitle        : "COVID-19 related social media use and well-being"

@@

abstract          : "In times of crisis such as the COVID-19 pandemic citizens need to stay informed about recent events, political decisions, or mandatory protection measures. To this end, many people use various types of media, and increasingly social media. However, because social media are particularly engaging, some find it hard to disconnect. In this preregistered study, I investigated whether using social media for COVID-19 related reasons affected psychological well-being. To answer this question I analyzed data from the Austrian Corona Panel Project, which consists of 3,485 participants. Well-being was measured at all 32 waves, and communication at six specific waves. I ran three random effects within between models, controlling for several stable and varying confounders. Results showed that the effects of COVID-19 related social media use on well-being were very small, arguably too small to matter. Fears that social media use during times of crisis critically impairs well-being are not supported."
keywords          : "COVID-19, well-being, social media, news use, panel study."

@@

header-includes:
  - |
    \makeatletter
    \renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-1em}%
      {\normalfont\normalsize\bfseries\typesectitle}}

    \renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-\z@\relax}%
      {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
    \makeatother
    \usepackage{endnotes}
    \setlength{\parskip}{0em}
    \raggedbottom
    \note{\clearpage}

csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass     : "apa7"
# output            :
#   papaja::apa6_pdf:
#     latex_engine  : xelatex
# output            : papaja::apa6_docx
output            : html_document
---

```{r analysis-preferences, include=F, cache=F}
set.seed(170819) # Seed for random number generation
knitr::opts_chunk$set(cache.extra = knitr::rand_seed,
                      cache=T, echo=F)
```

@@

# load packages
packages <- c("devtools", "english", "ggplot2", "gridExtra", "kableExtra", 
              "lavaan", "lme4", "lmerTest",
              "magrittr", "papaja", "semPlot", "tidyverse")
lapply(c(packages, "td"), library, character.only = TRUE)

# load workspace created from analyses.rmd to use results
load("data/workspace_1.RData")
```

@@

As a result, with this study I want to build on this research and investigate whether or not COVID-19 related social media use affected well-being during the pandemic.
To this end, I analyzed a large-scale panel study from the Austrian Corona Panel Project [@kittelAustrianCoronaPanel2020]. 
The panel consists of **32** waves and has an overall sample size of **`r nrow(d_wide) %>% prettyNum(big.mark = ",")` participants**.
<!-- At least 1,500 participants took part per wave, and the sample is representative of the Austrian population. -->
The panel study collected a large number of psychological and demographic variables.
<!-- Being able to control for many confounding third variables, both stable and varying, together with the longitudinal design, this creates a unique opportunity  -->
I explicitly aimed to investigate the causal effects of COVID-19 related social media use on well-being.

@@

The hierarchical taxonomy of computer-mediated communication differentiates six levels of how people engage with digital technology.
First, the device (e.g., smartphone); second, the type of application (e.g., social networking site); third, the branded application (e.g., Twitter); fourth, the feature (e.g., status post); fifth, the interaction (e.g., one-to-many); and sixth, the message (e.g., content) [@meierComputermediatedCommunicationSocial2020a].
Whereas the first four levels focus on the communication _channel_, the last two address the communication _type_.
**Differentiating different communication channels and communication types is important, because the effects on well-being likely differ across communication channels and communication types.**
**Whereas active social media use such as chatting is routinely linked to improved well-being, passive use such as reading is considered more negative [@dienlinImpactDigitalTechnology2020].**
Similarly, branded apps are separate entities with potentially divergent effects. 
For example, @waterlooNormsOnlineExpressions2018 found that it's more adequate to express negative emotions on WhatsApp than on Twitter or on Instagram. 
**Especially during a pandemic, it makes sense to analyze if users engage with COVID-19 related content on Instagram, where communication is more positive, or on Facebook, where communication is more critical.**

**In this study, to measure the effects of social media use focused on COVID-19 related news and topics, I adopt both the channel and the type of communication perspective.**
**Together, this should offer a nuanced and comprehensive understanding of communication.**
First, I investigate how well-being is affected by different types of communication affect, namely active and passive use.
Defining what constitutes active and what passive use is not always clear, and different understandings are currently discussed [@ellisonWhyWeDon2020; @meierComputermediatedCommunicationSocial2020a].
Reading is generally considered as passive and writing as active, while there are also specific behaviors such as such liking or sharing content that fall somewhere in-between [@meierDoesPassiveSocial2022].
In this study, I hence distinguish (a) _reading_ (passive), (b) _posting_ (active), and (c) _liking and sharing_ COVID-19 related posts (both active and passive).
Second, I analyze how using the most prominent branded applications affects well-being, and whether this effect changes across applications.
The branded applications investigated here are Facebook, Twitter, Instagram, WhatsApp, and YouTube---which were, at the time of writing, the most relevant social media apps.

**Worth noting, this study is not about _general_ social media use during times of COVID, but on social media use _focused_ on COVID-19 related content.** 
**For example, posting thoughts about the pandemic, reading posts and comments, or retweeting COVID-19 related news.**

@@

Can media use be such a factor?
In advance, there doesn't seem to be a clear winner, and it's likely that both positive and negative effects cancel each other out.
Empirically, social media use on average does not have a strong effect on well-being [@meierComputermediatedCommunicationSocial2020a].
According to the Differential Susceptibility to Media Effects Model [@valkenburgDifferentialSusceptibilityMedia2013], the effects of media use differ across individuals **and content**.
Whereas for some media are beneficial, for others they are harmful.
**Whereas some content provides opportunities (education, advice), other content creates risks (misinformation, hate). [@livingstoneEuropeanResearchChildren2018].**
Social media can impair well-being when causing embarrassment, stress, or disinformation, and they can improve well-being when providing connectedness, information, or entertainment [@buchiDigitalWellbeingTheory2021].
On average, however, effects are often small or negligible.

**Two of the most prominent media effect theories argue (indirectly) against strong average negative impacts.**
**According to mood management theory [@zillmannMoodManagementCommunication1988], using media substantially affects people's moods.**
**Effects can be stimulating or overwhelming, relaxing or boring.**
**After some time, users implicitly learn what media help them balance their mood and affect according to their own situational needs [@zillmannMoodManagementCommunication1988].**
**Those media that eventually become part of one's media repertoire are hence, on average, beneficial for users and their moods.**
**Using experience sampling of well-being and logs of social media use, a study with 82 participants from Italy found that after episodes of social media use, levels of positive affect increased significantly [@marcianoDynamicsAdolescentsSmartphone2022].**

**While mood management theory considers media use mainly driven by implicit learning experiences, uses and gratifications theory upholds that the process is more explicit and rational [@katzUsesGratificationsResearch1973].**
**Users select those media that they expect to have a desired effect, for example on mood, knowledge, or entertainment.** 
**If those beneficial media effects do not exist or if they are not expected, people will spend their time elsewhere.**
**And social media offer several beneficial effects, explaining why they are used that much.**
**They help find relevant information, maintain and foster relationships, express one's personality, and entertain oneself [@pelletierOneSizeDoesn2020].**
**In conclusion, because people spend so much time on social media consuming COVID-19 related content, according to both mood management theory and uses and gratifications theory we wouldn't expect to find strong average negative effects.**

@@

**Because media effects can differ across users, situations, and content [@livingstoneEuropeanResearchChildren2018; @valkenburgDifferentialSusceptibilityMedia2013], I now briefly focus on the effects of COVID-19 related social media use specifically.**
First, one could assume a _direct_ negative effect on well-being, and especially on positive or negative affect, which are more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Consuming such news can be depressing [@dornemannHowGoodBad2021], perhaps even changing some general perspectives on life.
<!-- So, just like being hit by a hammer hurts and we don't need any "mediating mechanism", this could be the case here as well. -->
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects [@dornemannHowGoodBad2021].
A study with 2.057 respondents from Italy reported that during the pandemic virtual community and social connectedness even increased during the pandemic [@guazziniSecondWaveAnalysis2022].
**In Finland, in a sample of 735 participants, levels of loneliness did not decrease during the pandemic, and people who engaged more on social media showed less loneliness [@latikkaLonelinessPsychologicalDistress2022].**
<!-- However, in light of a worldwide pandemic with millions of deaths, the negative direct effect seems more plausible. -->

There could also be _indirect_ effects.
When browsing social media for Covid-19 related news, many users reported being captivated to such an extent that they could not stop using social media [@kleinDarklySoothingCompulsion2021].
During the pandemic social media use was at an all-time high in the US [@statistaAverageDailyTime2021].
Although it is most likely that moderate social media use is not detrimental [@orbenTeenagersScreensSocial2020], overuse, however, might be more critical, and several studies have shown more pronounced negative effects for extreme users [@przybylskiLargescaleTestGoldilocks2017].
To explain, overuse could impair well-being if it replaces meaningful or functional activities such as meeting others, working, actively relaxing, or exercising.
**Another potentially negative mechanism at play are problematic social comparison processes. During the pandemic, several users shared how they successfully dealt with challenges such as physical distancing. In a study with 1131 residents from Wuhan in China [@yuePassiveSocialMedia2022], people who spent more time in quarantine also spent more time on social media. Those, who spent more time on social media also engaged in more upward social comparison, which was related to increased levels of stress.**
<!-- Therefore, if a society collectively overuses social media during a pandemic, there is potential for negative effects. -->
<!-- If overuse replaces such activities it's reasonable to assume that it's also detrimental. -->

On the other hand, one can make the case that using social media for COVID-19 related reasons might even be beneficial, especially in times of a pandemic.
Exchanging COVID-19 related messages with friends via WhatsApp might replace the in-person contact one would have otherwise, but which is literally impossible at the time.
In situations where meaningful and functional activities are prohibited, using social media to exchange about COVID-19 related topics might not be the worst idea.
Besides, given that nowadays a large number of experts, scientists, and politicians converse directly on social media, one can get first-hand high quality information on current developments.
<!-- On the other hand, there is of course also much disinformation, and "bingeing" on COVID-19 fake news might also pose risks for impaired well-being. -->

@@

In conclusion, in light of the theoretical considerations and empirical studies presented above, I expect that COVID-19 related communication on social media doesn't affect well-being in a meaningful or relevant way.
**This general hypothesis will be analyzed specifically for the communication types of (a) time spent reading, (b) liking and sharing, and (c) actively posting COVID-19 related content.**
**In addition, I will analyze how well-being is influenced by spending time on five prominent social media apps, including (a) Facebook, (b) Instagram, (c) Twitter, (d) WhatsApp, and (e) YouTube.** 
**Three different well-being indicators will be differentiated: life satisfaction, positive affect, and negative affect.**

@@

The hypothesis explicitly states a causal effect.
In non-experimental studies, longitudinal designs can help investigate causality.
Using longitudinal designs alone, however, is not sufficient for establishing correct causal statements [@rohrerTheseAreNot2021].
In addition, we for example also need to control for confounding third variables.
**Importantly, when analyzing longitudinal (within-person) relationships and effects, it is important to control for _varying_ third variables.**
**Non-varying third variables only help control non-varying (between-person) relations.**

@@

Thus, only when controlling for _all_ relevant confounders, we can correctly estimate causality without bias [@rohrerThinkingClearlyCorrelations2018].
Obviously, we can never be entirely sure to have included all confounders, which makes absolute statements regarding causality virtually impossible.
In addition, when determining the overall causal effect, we need to make sure _not_ to control for mediating variables [@rohrerThinkingClearlyCorrelations2018], for doing so would bias our assessment of the causal effect.
Complicating matters further, it is often unclear if a variable is a mediator or a confounder.[^collider]
However, despite all these caveats, when controlling for relevant variables (that aren't mediators), we can be much more certain that we measured causality correctly.
The aim should therefore be to collect as many varying and non-varying confounders as possible (which I believe is seldom done in our field), while knowing that absolute certainty regarding causality cannot be reached.

@@

[^explore]: The data-set includes many other variables that one could also potentially control for, and I invite interested readers to download the and explore potential interesting relationships.

@@

Finally, one precondition of causality is temporal order. 
The cause needs to precede the effect.
Finding the right interval between cause and effect is crucial.
For example, if we want to understand the effect of alcohol consumption on driving performance, it makes a big difference if driving performance is measured one minute, one hour, one day, or one week after consumption.
<!-- Finding the right interval is difficult. -->
If variables are stable, longer intervals are needed; if they fluctuate, shorter intervals.
In the case of well-being, we need shorter intervals for **the more fluctuating** positive and negative affect and longer ones for **the more stable** life satisfaction [@dienlinImpactDigitalTechnology2020].
**Using social media can have instant effects on mood [@marcianoDynamicsAdolescentsSmartphone2022].**
**Effects on life satisfaction often take longer to manifest, for example because media use leads to actual changes in specific behaviors, which then affect life satisfaction [@dienlinDisplacementReinforcementReciprocity2017].**
Choosing the right interval is challenging, because especially short intervals are hard to implement in practice, often requiring advanced methods such as experience sampling (also known as in situ measurement or ambulant assessment) [@schnauber-stockmannMobileDevicesTools2020].
In this study, I hence analyze how using social media during the last week affected positive and negative affect during the same week.
In other words, if people during the last week engaged in more COVID-19 related social media use than the usually do, did they feel better or worse during that week than they usually do?
Regarding life satisfaction, I implemented a longer interval.
If people during the last week used COVID-19 related social media more than the usually do, were they at the end of the week more or less satisfied with their lives than they usually are?
I hence analyze if when a person changes their social media diet, are there (a) _simultaneous_ changes in their affect and (b) _subsequent_ changes in their life satisfaction?
**These relations will be controlled for varying confounders, which fosters a causal interpretation.**
Similar approaches were implemented by other studies [@johannesNoEffectDifferent2022; @scharkowHowSocialNetwork2020], and they are considered a best practice approach toward analyzing causality [@bellFixedRandomEffects2019].

@@

**The data come from the Austrian Corona Panel Project [@kittelAustrianCoronaPanel2021], which is a large-scale standalone panel study. **
**The data are hosted on AUSSDA and are publicly available here: https://doi.org/10.11587/28KQNS.**
**At the time of writing, the official website included a data-set consisting of 24 waves. **
**For the analyses presented here, I was able to receive an advance version consisting of all 32 waves.**
**The study was conducted between March 2020 and June 2022, and data collection is now officially finished.**
**It contains 32 waves.**
Between March 2020 and July 2020, the intervals were weekly, and afterward the intervals were monthly.
Each wave consists of at least 1,500 respondents.
The overall sample size was _N_ = `r d_raw %>% nrow() %>% prettyNum(big.mark = ",")`, and `r d_long %>% nrow() %>% prettyNum(big.mark = ",")` observations were collected.
Panel mortality was compensated through a continuous acquisition of new participants. 
All respondents needed to have access to the internet (via computer or mobile devices such as smartphones or tablets).
They were sampled from a pre-existing online access panel provided by the company Marketagent, Austria.
Respondents were asked and incentivized with 180 credit points to participate in each wave of the panel.

@@

The factorial validity of the scales were tested with confirmatory factor analyses (CFA).
Because Mardia’s test showed that the assumption of multivariate normality was violated, I used the more robust Satorra-Bentler scaled and mean-adjusted test statistic (MLM) as estimator.
To avoid over-fitting, I tested the scales on more liberal fit criteria (CFI > .90, TLI > .90, RMSEA <. .10, SRMR < .10) [@klinePrinciplesPracticeStructural2016].
<!-- Finally, REWB-models cannot model latent variables. -->
<!-- To increase precision, I therefore exported factor scores from the CFAs for positive and negative affect. -->
Mean scores were used for positive and negative affect.
<!-- Respondents who answered less than 50% of all questions were removed. -->
**Missing responses were imputed using multiple imputation with predictive mean matching (five iterations, five data-sets).**
**Predictive mean matching allowed to impute also categorical variables.**
**All variables were imputed except the media use measures, as they were not collected on each wave.**
**All variables included in the analyses presented here were used to impute missing data.**
**For the main analyses, results were pooled across all five data-sets.**

@@

Social media use was measured for all participants on waves 1, 2, 8, 17, 23, **and 28**.
Freshly recruited respondents always answered all questions on social media use.

@@

First, when looking at the variables from a descriptive perspective (Figure \@ref(fig:fig-descriptives)), we see that all well-being measures did not change substantially across the different waves of data collection.
COVID-19 related media use, however, decreased slightly at the beginning of the study and remained stable after approximately six waves.
The initial decrease might be explained by the fact that the collection of data began at the end of March 2020, hence approximately three months after the pandemic began.
It could be that after an initial uptick, COVID-19 related social media use was already declining at the time, returning to more normal levels.

@@

The study's main hypothesis was that the effects of social media use on well-being would be trivial.
Regarding the effects of different communication _types_---that is, reading vs. sharing vs. posting---all within-person effects fell completely within the a-priori defined null region (see Figure \@ref(fig:fig-between)).
For example, respondents who used social media more frequently than usual to read about COVID-19 related topics did not show a simultaneous change in life satisfaction (_b_ = `r dat_fig_results_within %>% filter(dv == "Life satisfaction" & iv == "Reading" & type == "within" & Analysis == "Regular") %>% select(estimate) %>% round(2)` [95% CI `r dat_fig_results_within %>% filter(dv == "Life satisfaction" & iv == "Reading" & type == "within" & Analysis == "Regular") %>% select(conf.low) %>% round(2)`, `r dat_fig_results_within %>% filter(dv == "Life satisfaction" & iv == "Reading" & type == "within" & Analysis == "Regular") %>% select(conf.high) %>% round(2)`]).
As a result, the hypothesis was supported for all COVID-19 related types of social media communication.
**However, two effects were statistically significantly different from zero.**
**Users who wrote more COVID-19 related posts than usual were also slightly less satisfied with their lives as usual (_b_ = `r dat_fig_results_within %>% filter(dv == "Life satisfaction" & iv == "Posting" & type == "within" & Analysis == "Regular") %>% select(estimate) %>% round(2)` [95% CI `r dat_fig_results_within %>% filter(dv == "Life satisfaction" & iv == "Posting" & type == "within" & Analysis == "Regular") %>% select(conf.low) %>% round(2)`, `r dat_fig_results_within %>% filter(dv == "Life satisfaction" & iv == "Posting" & type == "within" & Analysis == "Regular") %>% select(conf.high) %>% round(2)`]).**
**Users who wrote more COVID-19 related posts than usual also experienced slightly more negative affect than usual (_b_ = `r dat_fig_results_within %>% filter(dv == "Negative affect" & iv == "Posting" & type == "within" & Analysis == "Regular") %>% select(estimate) %>% round(2)` [95% CI `r dat_fig_results_within %>% filter(dv == "Negative affect" & iv == "Posting" & type == "within" & Analysis == "Regular") %>% select(conf.low) %>% round(2)`, `r dat_fig_results_within %>% filter(dv == "Negative affect" & iv == "Posting" & type == "within" & Analysis == "Regular") %>% select(conf.high) %>% round(2)`]).**
**There was a small and statistically nonsignificant trend that reading COVID-19 related content slightly increased life satisfaction (_b_ = `r dat_fig_results_within %>% filter(dv == "Life satisfaction" & iv == "Reading" & type == "within" & Analysis == "Regular") %>% select(estimate) %>% round(2)` [95% CI `r dat_fig_results_within %>% filter(dv == "Life satisfaction" & iv == "Reading" & type == "within" & Analysis == "Regular") %>% select(conf.low) %>% round(2)`, `r dat_fig_results_within %>% filter(dv == "Life satisfaction" & iv == "Reading" & type == "within" & Analysis == "Regular") %>% select(conf.high) %>% round(2)`], _p_ = `r dat_fig_results_within %>% filter(dv == "Life satisfaction" & iv == "Reading" & type == "within" & Analysis == "Regular") %>% select(p.value) %>% my_round("p")`).**
**At the same time, there was also a small and statistically nonsignificant trend that reading COVID-19 related content decreased positive affect (_b_ = `r dat_fig_results_within %>% filter(dv == "Positive affect" & iv == "Reading" & type == "within" & Analysis == "Regular") %>% select(estimate) %>% round(2)` [95% CI `r dat_fig_results_within %>% filter(dv == "Positive affect" & iv == "Reading" & type == "within" & Analysis == "Regular") %>% select(conf.low) %>% round(2)`, `r dat_fig_results_within %>% filter(dv == "Positive affect" & iv == "Reading" & type == "within" & Analysis == "Regular") %>% select(conf.high) %>% round(2)`], _p_ = `r dat_fig_results_within %>% filter(dv == "Positive affect" & iv == "Reading" & type == "within" & Analysis == "Regular") %>% select(p.value) %>% my_round("p")`).**
